---
title: Recidivism and Staffing Requirements for Programs at GDOC
subtitle: Report
author: "GDOC"
nocite: |
  @*
format:
  pdf:
    toc: true
    toc-depth: 2
    number-sections: true
    number-depth: 2
    output-file: "finalreportGDOC"
    output-ext:  "pdf"
editor: source
---
  
```{r}
#| echo: false
#| eval: true
process_qmd <- function(file, fpath_in = "images/", fpath_out = "images/") {
  doc <- readLines(file)
  end_yaml <- which(doc == "---")[2]
  out_doc <- doc[seq(end_yaml+1, length(doc))]
  if (fpath_in != fpath_out) {
    out_doc <- stringr::str_replace_all(out_doc, fpath_in, fpath_out)
  }
  res <- knitr::knit_child(text = out_doc, quiet = TRUE, options = list(eval = FALSE, echo = TRUE))
  return(res)
}
```

This report includes the one-year recidivism rates for evidence-based programming provided in the Gotham Department of Corrections facilities for high-risk people. Staffing numbers describe how many full-time employees are needed over the course of a year to keep each program running. Each program facilitator completes extensive training and must complete eight hours of continuing education each year. 

```{r}
#| output: asis
#| echo: false
#| eval: true
#| message: false
idx <- process_qmd("finaltable.qmd")
cat(unlist(idx), sep = '\n')
```

```{r}
#| output: asis
#| echo: false
#| eval: true
#| message: false
idx <- process_qmd("finalreport.qmd")
cat(unlist(idx), sep = '\n')
```